---
title: "What risyphus can do for you"
author: "Fridtjof Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What risyphus can do for you}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style


## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

---

Tools/utility functions to facilitate recurrent tasks with focus on
logistic regression, Cox regression, and studies/clinical trials with two groups.

This package contains functions that facilitate laborious steps
in most data analysis tasks, such as basic variable description,
compilation of baseline tables comparing two groups, odds-ratio tables
based on logistic regression, or hazard-ratio tables based on
Cox proportional hazards models. The name of this package is derived
from the Greek mythological figure Sisyphus (Sisyphus for R = risyphus),
who was doomed to roll a big boulder up a hill just to see it roll down
again ad infinitum. Somehow I related to the laborious task of keeping all
standard tables up to date in a developing project as Sisyphean, and it all
too often consumed too much time and energy devoted to a project. My
intention with this package is to do standard tasks well and to free time
and resources for better statistical analyses and to facilitate
reproducible computing.

---


## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
