---
title: "How to use risyphus: A simple example"
author: "Fridtjof Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use risyphus: A simple example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



---



## Example Description
In this simple example, we will use a build-in data set to create a table
that compares two groups with each other. In the medical field, such tables are
almost always presented for

* clinical trials, in which case they show demographics, anthropomorphics, medical
history, etc. at baseline, i.e. at enrollment/before the trial's active intervention
starts. 
* observational studies, in which case they typically show demographics,
anthropomorphics, medical history, etc. at the point in time that defines "entry
into the study" such as the day a patient presented with certain symptoms.

Remark: In a randomized clinical trial (RCT) associated p-values are debatable
since all "imbalances" must be due to coincidental differences and cannot be associated
with treatment assignment. Nevertheless, p-values are typically provided and can
be used to assess the integrity of the random assignment to the study arms and/or
whether specific traits are effectively balanced across study arms.

## A baseline table created with risyphus
The synthetic data set `example_data` is included in the installation of risyphus. 
It includes 100 (simulated) individuals and 5 variables:

1. ID: Fictitious ID
1. Group: group membership - GroupA or GroupB (as text)
1. Gender: Fictitious gender as example of a binary variable (as text)
1. Age: Fictitious age in years as example of a continuous variable
1. Some.measure: Fictitious continuous measure with skewed distribution.
1. Educ: Fictitious education level as example of a factor (with 8 possibly unused factor levels)

It can be printed by simply calling it:
```{r }
library(risyphus)
summary(example_data)
```

The information about which variables to include in the table and which
statistical tests to apply is provided in a second data set named `example_variables`:
```{r }
example_variables

```

A variable *not* listed above is the one that defines the two groups, namely
"Group" in our example.  This information needs to be submitted to the 
function `BLtable` as well, and we get a compiled "baseline table" as
follows (depending on the width of the output, the rectangular character matrix
might be displayed as several blocks):
```{r }
BLtable(data = example_data, info = example_variables, group.var = "Group")

```

The output is a character matrix and is intended to be submitted to some other
function that creates the table to be displayed. The `knitr`-function `kable`
is a simple choice (`knitr` must be installed for this to work):
```{r kable}
mytable <- BLtable(data = example_data, info = example_variables, group.var = "Group")
knitr::kable(mytable,
             caption = "Table: Baseline covariates by gender. Shown are
          mean (std.dev.) and count (%), respectively, depending on
          whether the variable is continuous or discrete.")

```

Other simple functions that can be used to display the table in html or other
formats are provided by the
[pander package](https://cran.r-project.org/web/packages/pander/pander.pdf)
and the
[xtable package](https://cran.r-project.org/web/packages/xtable/xtable.pdf).
The [htmlTable package](https://cran.r-project.org/web/packages/htmlTable/htmlTable.pdf)
gives detailed control over tables in html. 

## Anatomy of the Function Call
The function call above
```{r eval=FALSE}
BLtable(data = example_data, info = example_variables, group.var = "Group")
```
provides information about the data set (example_data), which variables to use 
for the table (example_variables), and which variable defines the two groups ("Group").


### `group.var = "Group"`
"Group" *must* be a binary variable and anything else than exactly two groups will
cause an error. If "Group" is not binary, the data set *must* be processed to make it
binary *before* it is submitted to `BLtable`. Regularly occurring problems in
"perceived binary" variables are coding errors leading to more than two groups,
which can be checked by:
```{r }
unique(example_data$Group)

```

or by
```{r }
levels(example_data$Group)
```

or by printing a summary, which also allows to verify that none of the two groups
are empty (if `summary` does not print a count-summary as below, the variable is
not a formal factor and will need to be changed accordingly):
```{r }
summary(example_data$Group)
```



Another source of "unexpectedly" more than two groups originates from
subsetting a factor without dropping unused factor levels (the default in R is
to *not* drop factor levels). E.g., if race is coded as "Black", "White", "Other/Multiple",
and the subset of "Black" and "White" is submitted to `BLtable`, the factor (Race)
will still have three (and not two!) levels, unless the user took action to drop the
unused factor level. Garrett Grolemund and Hadley Wickham describe 
[factors](https://r4ds.had.co.nz/factors.html) in detail in
their book [R for Data Science](https://r4ds.had.co.nz/factors.html).

### `info = example_variables`
This data set provides information about which variables to include in the table,
which name to give/print in the table, and which statistical test to apply
for that variable. Tests for categorical variables *require* the variable to be 
a (formal) factor in the data set. Garrett Grolemund and Hadley Wickham describe 
[factors](https://r4ds.had.co.nz/factors.html) in detail in
their book [R for Data Science](https://r4ds.had.co.nz/factors.html).

While a name different from the variable name can be specified to be printed
in the table, all *factor levels* will have the name extracted from their
respective factor label, and the order of display will be the order defined 
by the formal factor definition in the data set. You can change that order by
first reordering the respective factor levels and then submitting the data
to `BLtable`. Variable names are more restrictive, which motivates the option
to allow for a different name.  Regularly it will be helpful to give a more
descriptive name than the formal variable name, and/or to add additional
information such as a measurement unit. A good table should prominently
display all information needed to correctly interpret its content. Obscure variable
names, unclear measurement units, and/or undefined grouping levels should be
avoided.

In practical use, it is recommended to organize `info = example_variables` in a
csv-file or spreadsheet and to import and pre-process it from there. That way 
hundreds of variables in potentially different versions can be organized in a single
spreadsheet and different columns define which variable is used in which table 
(frequently very helpful if a complete analysis contains many tables and 
subgroup analyses).

The order of display of the variables is defined by the order of appearance in
`info = example_variables`. It is regularly beneficial to logically group
variables, typically first listing demographics, anthropomorphics, and
socioeconomic information, followed by medical history, etc. 

### `data = example_data`
The submitted data set *must* contain all variables used as specified in 
`info = example_variables` and the variable `group.var = "Group"`.  To reiterate:
All variables for categorical measures *must* be formal factors in the data set, and
the variable defining the two groups *must* be a factor with exactly two levels.
Garrett Grolemund and Hadley Wickham describe 
[factors](https://r4ds.had.co.nz/factors.html) in detail in
their book [R for Data Science](https://r4ds.had.co.nz/factors.html).

### Factors in R
Did I mention that [factors](https://r4ds.had.co.nz/factors.html) are described
in detail in the book [R for Data Science](https://r4ds.had.co.nz/factors.html)?
You can use any other book or source of information to learn about factors in R,
of course, but you really need to be on top of the "factor game" if you want
to use risyphus effectively. Especially, you will need to know how to 
do the following:

* create factors in R,
* change factor labels,
* drop unused factor levels,
* reorder factor levels.

## The Power of risyphus
Let's use the above outlined principles to produce a summary table that
contains the following changes:

* use two digits precision even for proportions.
* display four digits for the p-values.

Here is the code to produce such a table (note that the needed changes are 
limited to two options in the `BLtable` function):

```{r kable-digits}
mytable <- BLtable(data = example_data, info = example_variables, group.var = "Group",
                   sign.digits.prop = 2, pvalue.digits = 4)
knitr::kable(mytable,
             caption = "Table: Baseline covariates by gender. Shown are
          mean (std.dev.) and count (%), respectively, depending on
          whether the variable is continuous or discrete.")

```

And how about two tables for each gender? Here is the code for females:

```{r kable-females}
this_data <- example_data[example_data$Gender == "female",]
this_info <- example_variables[example_variables$Variable != "Gender", ]
mytable <- BLtable(data = this_data, info = this_info, group.var = "Group")
knitr::kable(mytable,
             caption = "Table: Baseline covariates for females. Shown are
          mean (std.dev.) and count (%), respectively, depending on
         whether the variable is continuous or discrete.")
```


And here for males:

```{r kable-males}
this_data <- example_data[example_data$Gender == "male",]
this_info <- example_variables[example_variables$Variable != "Gender", ]
mytable <- BLtable(data = this_data, info = this_info, group.var = "Group")
knitr::kable(mytable,
             caption = "Table: Baseline covariates for males. Shown are
          mean (std.dev.) and count (%), respectively, depending on
         whether the variable is continuous or discrete.")
```

Note that above gender specific tables contain 63 numbers each, and the combined
table (both genders) contains 76 numbers for a total of 202 numbers for these
three short tables only!

Also note that that the function call would be *unchanged* when including
additional variables: all we have to do is to list these additional
variables in the data set submitted to `info = ` (assuming the variable exists 
already in the main data set).


## Other Functions in risyphus and further information
All other functions relating to tables in `risyphus` are working in the same
way as `BLtabel` shown above and follow this approach:

1. call the function of your choice to compile the table information,
1. pass the resulting character matrix to a different function that 
displays the output. 

The vignette *Introduction to risyphus* 
[`vignette("Introduction", package = "risyphus")`] contains a table with the
core functions available after installing risyphus.

Forthcoming: The vignette *Combining  risyphus and R Markdown* 
[`vignette("risyphus-RMarkdown", package = "risyphus")`] contains a detailed example
of how to harvest the true power of risyphus (and R Markdown) to generate reports.

Forthcoming: The vignette *risyphus sample code* 
[`vignette("risyphus-code", package = "risyphus")`] contains sample code
for each function in risyphus including a suitable subsequent call to 
a function that produces the corresponding html-table (including hints for
trouble shooting).



## Software/R packages used for this document

The following software/R-packages are used in this vignette and
are gratefully acknowledged:


| Package  | Reference       |
|:-----:|:---------------------------------------------------------------------|
| R | `r citation("base")$textVersion` |
| R Studio | RStudio Team (2016). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL http://www.rstudio.com/. |
| rmarkdown | `r citation("rmarkdown")$textVersion`
| knitr | `r citation("knitr")$textVersion` |


*****

End of document

*****
